---
title: "TidyTuesday Week 9 - US PhDs"
author: "James Stone"
date: "19 February 2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.height = 8)
library(tidyverse)
library(collapsibleTree)
extrafont::loadfonts(device="win")
```

Load Data

```{r}
phd_data <- read_csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2019/2019-02-19/phd_by_field.csv")
```

Check for missing data

```{r}
sum(is.na(phd_data$n_phds))
sum(phd_data$n_phds == 0, na.rm=T)
```

If there was no 0 values then would seem sensible to assume the NA values could be re-coded to 0. Best to just remove them in this case, zero by omission. 

# collapsibleTree

For this week's #TidyTuesday I'm going to look at the [collapsibleTree](https://cran.r-project.org/web/packages/collapsibleTree/index.html) package and learn the basics. By picking a new package/vis-technique each week to go with each #tidyTuesday dataset it presents the oportunity to both practice tidyverse wrangling and gain new knowledge each time.

## A basic collapsible tree diagram


```{r}
# Broad Field -> Year -> Total
phd_data %>% 
  filter(!is.na(n_phds)) %>% 
  group_by(year, broad_field) %>% 
  summarise(
    total = sum(n_phds)
  ) %>% 
  collapsibleTree(
    c("broad_field", "year", "total")
  )
```

## collapsibleTreeSummary()

Changing to **collapsibleTreeSummary()** instead of collapsibleTree changes the default colors and adds some basic hover information (number of leaf nodes):

```{r}
# Broad Field -> Year -> Total
phd_data %>% 
  filter(!is.na(n_phds)) %>% 
  group_by(year, broad_field) %>% 
  summarise(
    total = sum(n_phds)
  ) %>% 
  collapsibleTreeSummary(
    hierarchy = c("broad_field", "year"),
    zoomable = F
  )
```

## Order nodes by total PhDs awarded

We can add the attribute argument to use a different metric to the internal weighting which will define the order in which leaves are presented, the information in the hover, and things like gradient fill. We can take "total" out of the hierarchy now, and this could make sense as the extension from that node will only ever have one node leading to the total, we might as well get that information from the hover:

```{r}
# Broad Field -> Year -> Total
phd_data %>% 
  filter(!is.na(n_phds)) %>% 
  group_by(year, broad_field) %>% 
  summarise(
    total = sum(n_phds)
  ) %>% 
  collapsibleTreeSummary(
    hierarchy = c("broad_field", "year"),
    attribute = "total",
    maxPercent = 50,
    zoomable = F
  )
```

## map the size of displayed nodes to values from a variable

Use the **nodeSize** argument:

```{r}
# Broad Field -> Year -> Total
phd_data %>% 
  filter(!is.na(n_phds)) %>% 
  group_by(year, broad_field) %>% 
  summarise(
    total = sum(n_phds)
  ) %>% 
  collapsibleTreeSummary(
    hierarchy = c("broad_field", "year"),
    attribute = "total",
    nodeSize = "total",
    zoomable = F
  )
```

## Custom color pallete

### collapsibleTree()

I wanted an element of gradient-fill within each level. From what I could gather the only way to do this was to carefully construct a vector of colours to fill based on the hierarchy I had specified which I did via the following method (the match between the second/third level main colour could be better if I had more time!):

```{r}
phd_data %>% 
  filter(!is.na(n_phds)) %>% 
  group_by(year, broad_field) %>% 
  summarise(
    total = sum(n_phds)
  ) %>% 
  collapsibleTree(
    hierarchy = c("broad_field", "year", "total"),
    fill = 
      c(
        "steelblue4",
        RColorBrewer::brewer.pal(n = length(unique(.$broad_field)), "Accent"),
        unlist(purrr::map(
          c("Greens","Purples","Oranges","Greys","Blues","Reds","Blues"),
          function(x) colorRampPalette(RColorBrewer::brewer.pal(n = 9, name = x))(length(unique(.$year)))
        )),
        rep("#43a3ca",70)
      ),
    zoomable=F
  )
```

## collapsibleTreeSummary()

The way colouring works is different in the **summary()** wrapper. I couldn't quickly work out a way to do the same within-level gradient fill as previously. Any tips on how to do that would be great.

```{r}
phd_data %>% 
  filter(!is.na(n_phds)) %>% 
  group_by(year, broad_field) %>% 
  summarise(
    total = sum(n_phds)
  ) %>% 
  collapsibleTreeSummary(
    hierarchy = c("broad_field", "year"),
    attribute = "total",
    fillFun = colorspace::heat_hcl,
    percentOfParent = TRUE,
    zoomable=F,
    nodeSize = "total"
  )
```

# Conclusion

Perhaps not the most sensible way to visualise the data here but an interesting visualisation tool to know about and have 'in the locker' for other problems down the line. Particularly useful when the volume of data needed to visualise is particularly large and a more consise alternative doesn't exist given how easy it is to produce the collapsible elements. 
